some title

### 2018-01-06 2018/paper.tex
rss ess tss

\begin{figure}[ht!]
\begin{center}
\includegraphics{images/corr_def.pdf}
\caption{Random vector $X$ of length $\sigma_X$ and random vector $Y$ of length $\sigma_Y$, $\cos \varphi$ is the angle between $X$ and $Y$.}
\label{fig:corr_def}
\end{center}
\end{figure}

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics{images/corr_proj.pdf}
\caption{Random vector $Y$ projected onto random vector $X$.}
\label{fig:corr_proj}
\end{center}
\end{figure}

@@

\begin{theorem}
A linear regression model with one explanatory variable and constant term has the property
\[
\sCorr(y, \hat y) = sign(\hat \beta_2) \sCorr(y, x)
\]
\end{theorem}

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_xy_centered.pdf}
\label{fig:corr_pos_xyc}}
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_yhat.pdf}
\label{fig:corr_pos_yhatc}}
\caption{\subref{fig:corr_pos_xyc}: `Centred' $x$ and $y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$; \subref{fig:corr_pos_yhatc}: `Centred' $\hat y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$.}
\end{center}
\end{figure}

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics{images/corr_eq_negative.pdf}
\label{fig:corr_negative}
\caption{Case of $\beta_2 < 0$.}
\end{center}
\end{figure}
\end{proof}

>>>>

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_def.pdf}
\caption{Random vector $X$ of length $\sigma_X$ and random vector $Y$ of length $\sigma_Y$, $\cos \varphi$ is the angle between $X$ and $Y$.}
\label{fig:corr_def}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_proj.pdf}
\caption{Random vector $Y$ projected onto random vector $X$.}
\label{fig:corr_proj}
\end{center}
\end{figure}

@@

\begin{theorem}
A linear regression model with one explanatory variable and constant term
\[
y = \beta_1 + \beta_2 x + \varepsilon
\]
has the property
\[
\sCorr(y, \hat y) = sign(\hat \beta_2) \sCorr(y, x)
\]
\end{theorem}

@@

\begin{figure}[h!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_xy_centered.pdf}
\label{fig:corr_pos_xyc}}
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_yhat.pdf}
\label{fig:corr_pos_yhatc}}
\caption{\subref{fig:corr_pos_xyc}: `Centred' $x$ and $y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$; \subref{fig:corr_pos_yhatc}: `Centred' $\hat y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$.}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_eq_negative.pdf}
\label{fig:corr_negative}
\caption{Case of $\beta_2 < 0$.}
\end{center}
\end{figure}
\end{proof}

\subsection{RSS + ESS = TSS}

\marginnote{
Consider a regresion model with $n$ observations and $k$ explanatory variables including a constant unit vector
\[
y = X \beta + \varepsilon
\]
The OLS estimator for the vector of coefficients $\beta$ is
\[
\hat \beta = (X^T X)^{-1} X^T y
\]
and the residual vector is
\begin{align*}
\hat e &= y - \hat y \\
&= y - X \hat \beta \\
&= y - X (X^T X)^{-1} X^T y
\end{align*}
Then we define residual sum of squares (RSS), explained sum of squares (ESS) and total sum of squares (TSS) as follows:
\begin{align*}
RSS &= \lVert y - \hat y \rVert^2_2 \\
ESS &= \lVert \hat y - \bar y \rVert^2_2 \\
TSS &= \lVert y - \bar y \rVert \\
\end{align*}
Disclosing parentheses and using the fact that $\hat{y}^T y = \hat{y}^T \hat{y}$
\begin{align*}
\hat{y}^T y &= \beta^T X^T y \\
&=  y^T X (X^T X)^{-1} X^T y \\
\hat{y}^T \hat{y} &= \beta^T X^T X \beta \\
&= y^T X (X^T X)^{-1} X^T X (X^T X)^{-1} X^T y \\
&= y^T X (X^T X)^{-1} X^T y
\end{align*}
we obtain
\begin{align*}
RSS &= y^T y -\hat{y}^T \hat{y} \\
ESS &= \hat{y}^T \hat{y} - \hat{y}^T \bar{y} + \bar{y}^T \bar{y} \\
TSS &= y^T y - 2 y^T \bar y +  \bar{y}^T \bar{y}
\end{align*}
When putting everything together all the terms cancel out which proves
\[
ESS + RSS = TSS
\]
}

\begin{theorem}
A linear regression model with $n$ observations and $k$ explanatory variables including a constant unit vector
\[
y = X \beta + \varepsilon
\]
has the following property
\[
RSS + ESS = TSS
\]
where $RSS = \lVert y - \hat y \rVert^2_2$, $ESS = \lVert \hat y - \bar y \rVert^2_2$, $TSS = \lVert y - \bar y \rVert$.
\end{theorem}

\begin{proof}
The proof will be presented for the case of two regressor $x$ and $\mathbf{1}$ in order for the picture to be clear.
However, the same logic applies for the case of $k$ regressors.

We start with depicting the vectors $y \in \mathbb{R}^{n-2}$ and $x, \mathbf{1} \in \mathbb{R}^2$.
Then we project $y$ onto $Lin(x, \mathbf{1})$ and obtain $\hat y$ which is shown in Figure~\ref{fig:tss_base}.

From this picture we can immediately derive $\sqrt{RSS}$ as by definition this is the squared difference between $y$ and $\hat y$.

\begin{figure}[h!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/tss_sum_base.pdf}
\label{fig:tss_base} }
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/tss_sum_rss.pdf}
\label{fig:tss_rss}}
\caption{\subref{fig:tss_base}: Vectors $y \in \mathbb{R}^{n-2}$ and $\hat y \in Lin(x, \mathbf{1})$; \subref{fig:tss_rss}: Residual sum of squares.}
\end{center}
\end{figure}

So as to visualize $ESS$ and $TSS$ we first need to visualize vector of averages $\bar y$.
Geometrically this means projecting a vector onto a line spanned by vector $\mathbf{1}$.
It can be shown that the matrix corresponding to projecting onto the line spanned by a vector of all ones has the following form
\[
\frac{\mathbf{1}^T \mathbf{1}}{\mathbf{1} \mathbf{1}^T} = \frac{\begin{pmatrix} 1 & \ldots & 1 \end{pmatrix} \begin{pmatrix} 1 \\ \ldots \\ 1 \end{pmatrix}}{\sum_{i=1}^n 1} = \begin{pmatrix} \frac{1}{n} & \ldots & \frac{1}{n} \\ \vdots & \ddots & \vdots \\ \frac{1}{n} & \ldots & \frac{1}{n} \end{pmatrix}
\]

Now we both project $y$ and $\hat y$ onto $\mathbf{1}$ and following the definition obtain $\sqrt{TSS}$ as the difference vector $y - \bar y$ and $\sqrt{ESS}$ as the vector $\hat y - \bar y$.

\begin{figure}[h!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/tss_sum_tss.pdf}
\label{fig:tss_tss}}
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/tss_sum_ess.pdf}
\label{fig:tss_ess}}
\caption{\subref{fig:tss_tss}: Total sum of squares; \subref{fig:tss_ess}: Explained sum of squares.}
\end{center}
\end{figure}

The final step is to put everything together.
Note that since $y - \hat y$ is perpendicular to $lin(x, \mathbf{1})$ it is also perpendicular to $\hat y - \bar y$ and $\mathbf{1}$ as these vectoros are in $lin(x, \mathbf{1})$.
Then, applying the theorem of three perpendiculars we conclude that vecotr foot of vector $y - \bar y$ is the same point as the foot of the vector $\hat y - \bar y$.
Thus, we obtain a right angle triangle and can apply the Pythagorean theorem for the catheti $\sqrt{RSS}$ and $\sqrt{ESS}$ and the hypothenuse $\sqrt{TSS}$:
\[
\sqrt{RSS} + \sqrt{ESS} = \sqrt{TSS}
\]

\begin{figure}[h!]
\center{\includegraphics{images/tss_sum_final.pdf}}
\caption{$\sqrt{RSS} + \sqrt{ESS} = \sqrt{TSS}$}
\label{fig:tss_final}
\end{figure}

\subsection{Determination coefficient}
